---
title: "Model Development"
author: "Alyssa Willson"
date: "2023-04-26"
output: pdf_document
---

Put the variability in the climate drivers as separate drivers in GJAM
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
rm(list = ls())
library(rjags)
library(tidyverse)
library(sp)
library(rgeos)
```

## Load data

We will use a very small portion of our data for the last time period in our fossil pollen record. We'll load the data, subset it, and then remove the rest.

```{r}
#### Load data ####

load('FossilPollen/Data/full_melt_UMW.RData')

# For now let's join these to make things easier
pollen_data <- full_melt |>
  full_join(full_sd, by = c('time', 'long', 'lat', 'loc_time'))

load('Climate/processed_climate.RData')

#### Start processing climate drivers ####

climate <- clim_av |>
  mutate(Loc = paste0(Latitude, '_', Longitude),
         Loc_Year = paste0(Loc, '_', Year)) |>
  # Need to drop NA because we don't have climate drivers for some years of interest
  # So we won't model those years
  drop_na()

#### Match climate and fossil pollen data ####

# Take only one year to get unique locations
# All locations have data for all time points

spat_climate <- climate |> filter(Year == 1900)
# make spatial data object
coordinates(spat_climate) <- ~Longitude+Latitude

spat_pollen <- pollen_data |> filter(time == 1900)
coordinates(spat_pollen) <- ~long+lat

# Find distance between all points
d <- gDistance(spgeom1 = spat_pollen, spgeom2 = spat_climate, byid = T)
# Find closest climate data point to each pollen data point
mins <- apply(d, 2, which.min)
# result = 704 points corresponding to each grid cell of our pollen dataset

spat_pollen <- pollen_data |>
  filter(time == 1900) |>
  mutate(loc = paste0(lat,'_',long),
         match = mins)

spat_climate <- climate |>
  filter(Year == 1900)

full_spat_pollen <- pollen_data |>
  mutate(loc = paste0(lat,'_',long))

spat_map <- spat_pollen |>
  select(loc, match)

full_data <- matrix(, nrow = nrow(full_spat_pollen),
                    ncol = 33)

for(i in 1:nrow(full_spat_pollen)){
  loc <- full_spat_pollen$loc[i]
  match <- spat_map$match[which(spat_map$loc == loc)]
  year <- as.numeric(full_spat_pollen[i,1])
  full_data[i,1] <- match # Match between datasets
  full_data[i,2] <- loc # Location
  full_data[i,3] <- year # Year
  full_data[i,4:15] <- as.numeric(full_spat_pollen[i,2:13]) # ydata
  full_data[i,16:17] <- as.numeric(full_spat_pollen[i,14:15]) # Lat/lon of pollen
  full_data[i,18:29] <- as.numeric(full_spat_pollen[i,17:28]) # edata
  clim_loc <- spat_climate$Loc[match]
  temp <- climate |> filter(Loc == clim_loc) |> filter(Year == year)
  full_data[i,30] <- temp$Latitude # Lat of climate
  full_data[i,31] <- temp$Longitude # lon of climate
  full_data[i,32] <- temp$Temperature # temperature
  full_data[i,33] <- temp$Precipitation # precipitation
}

full_data <- as.data.frame(full_data)
colnames(full_data) <- c('Match', 'Location', 'Year', colnames(full_spat_pollen)[2:13],
                         'Pollen_Longitude', 'Pollen_Latitude', colnames(full_spat_pollen)[17:28],
                         'Climate_Latitude', 'Climate_Longitude', 'Temperature', 'Precipitation')

full_data <- full_data |>
  mutate(Year = as.numeric(Year),
         ash.x = as.numeric(ash.x),
         beech.x = as.numeric(beech.x),
         birch.x = as.numeric(birch.x),
         elm.x = as.numeric(elm.x),
         hemlock.x = as.numeric(hemlock.x),
         maple.x = as.numeric(maple.x),
         oak.x = as.numeric(oak.x),
         other_conifer.x = as.numeric(other_conifer.x),
         other_hardwood.x = as.numeric(other_hardwood.x),
         pine.x = as.numeric(pine.x),
         spruce.x = as.numeric(spruce.x),
         tamarack.x = as.numeric(tamarack.x),
         Pollen_Longitude = as.numeric(Pollen_Longitude),
         Pollen_Latitude = as.numeric(Pollen_Latitude),
         ash.y = as.numeric(ash.y),
         beech.y = as.numeric(beech.y),
         birch.y = as.numeric(birch.y),
         elm.y = as.numeric(elm.y),
         hemlock.y = as.numeric(hemlock.y),
         maple.y = as.numeric(maple.y),
         oak.y = as.numeric(oak.y),
         other_conifer.y = as.numeric(other_conifer.y),
         other_hardwood.y = as.numeric(other_hardwood.y),
         pine.y = as.numeric(pine.y),
         tamarack.y = as.numeric(tamarack.y),
         Climate_Latitude = as.numeric(Climate_Latitude),
         Climate_Longitude = as.numeric(Climate_Longitude),
         Temperature = as.numeric(Temperature),
         Precipitation = as.numeric(Precipitation))

cols <- colnames(full_data)
ydata_columns <- which(grepl('.x', cols, fixed = T))
xdata_columns <- which(cols == 'Temperature' | cols == 'Precipitation')

ydata <- full_data |>
  select(all_of(ydata_columns))
xdata <- full_data |>
  select(all_of(xdata_columns))

colnames(ydata) <- colnames(full_melt)[2:13]

ydata <- ydata |> rename(conifer = other_conifer,
                         hardwood = other_hardwood)
```

```{r}
clim_sub <- xdata |>
  slice_head(prop = 0.1)
data_sub <- ydata |>
  slice_head(prop = 0.1)

rm(clim, clim_av, climate, d, full_melt, full_sd, mins, pollen_data, spat_climate, spat_pollen, temp, xdata, ydata)
```

## Model 1: Observations drawn from a Gaussian distribution

Now, let's write a simple model to maek sure we know what we're doing in JAGS. We'll assume that our simulated data is normally distributed with mean $\mu$ and precision $\tau$:

$$y_{i,j} \sim \mathcal{N}(\mu, \tau)$$

$$\mu \sim \mathcal{N}(0, 0.001)$$

$$\tau \sim \text{Gamma}(0.001, 0.001)$$

```{r}
gauss_model <- "
model{
for(i in 1:nobs){
for(j in 1:ntaxa){
y[i,j] ~ dnorm(mu, tau)
}
}
mu ~ dnorm(0, 0.001)
tau ~ dgamma(0.001, 0.001)
}
"
```

Now, we'll compile and run our JAGS model.

```{r}
data <- list(nobs = nrow(data_sub),
             ntaxa = ncol(data_sub),
             y = data_sub)
jags.gauss_model <- jags.model(file = textConnection(gauss_model),
                               data = data,
                               n.chains = 3)
out.gauss_model <- coda.samples(model = jags.gauss_model,
                                variable.names = c('mu', 'tau'),
                                n.iter = 1000)
```

Now, we can do a quick visualization of this output. We expect to have converged, but we also expect that this won't be particularly informative since a Normal distribution is inappropriate for this data and each observation and taxon is fit separately.

```{r}
plot(out.gauss_model)
gelman.plot(out.gauss_model)
```

## Model 2: Observations drawn from a Dirichlet distribution

We can repeat this with a vrsion of the model in which the fractional composition observations are drawn from a Dirichlet distribution. THis is appropriate because fractional composition is limited to (0, 1) and is constrained to sum to 1. Using the same notation as before, we have

$$\boldsymbol{y}_i \sim \text{Dirichlet}(\boldsymbol{\alpha})$$

$$\alpha_j \sim \text{Gamma}(0.001, 0.001)$$

In this case, the taxa at each location are modeled jointly, as indicated by the vector $\boldsymbol{y}_i$. Each taxon has a different parameter describing the Dirichlet distribution, $\boldsymbol{\alpha}$, each with an uninformative Gamma prior, with support $(0, \infty)$.

```{r}
dirch_model <- "
model{
for(i in 1:nobs){
y[i,1:ntaxa] ~ ddirch(alpha[1:ntaxa])
}
for(j in 1:ntaxa){
alpha[j] ~ dgamma(0.001, 0.001)
}
}
"
```

```{r}
data <- list(nobs = nrow(data_sub),
             ntaxa = ncol(data_sub),
             y = data_sub)
jags.dirch_model <- jags.model(file = textConnection(dirch_model),
                               data = data,
                               n.chains = 3)
out.dirch_model <- coda.samples(model = jags.dirch_model,
                                variable.names = 'alpha',
                                n.iter = 1000)
```

```{r}
plot(out.dirch_model)
gelman.plot(out.dirch_model)
```

Convergence looks great. We will use this model as the basis for a model with climate drivers.

## Model 3: Including a hyperprior

Before modeling fractional composition with climate drivers, let's add a hyper prior to our model. When we add climate drivers to our model, we'll be incorporating a series of linear models with climate drivers and the $\alpha$ parameters of the Dirichlet distribution as the response. Therefore, we want to be able to manipulate the $\alpha$ parameters instead of specifying a vague prior for them.

We'll start with the following.

$$\boldsymbol{y}_i \sim \text{Dirichlet}(\boldsymbol{\alpha})$$

$$\alpha_j \sim \text{Gamma}(\alpha_{\alpha}, \beta_{\alpha})$$

$$\alpha_{\alpha} \sim \text{Gamma}(0.001, 0.001)$$

$$\beta_{\alpha} \sim \text{Gamma}(0.001, 0.001)$$

```{r}
dirch_hyper_model <- "
model{
for(i in 1:nobs){
y[i,1:ntaxa] ~ ddirch(alpha[1:ntaxa])
}
for(j in 1:ntaxa){
alpha[j] ~ dgamma(alpha_alpha, beta_alpha)
}
alpha_alpha ~ dgamma(0.001, 0.001)
beta_alpha ~ dgamma(0.001, 0.001)
}
"
```

```{r}
data <- list(nobs = nrow(data_sub),
             ntaxa = ncol(data_sub),
             y = data_sub)
jags.dirch_hyper_model <- jags.model(file = textConnection(dirch_hyper_model),
                                     data = data,
                                     n.chains = 3)
out.dirch_hyper_model <- coda.samples(model = jags.dirch_hyper_model,
                                      variable.names = c('alpha_alpha', 'beta_alpha'),
                                      n.iter = 1000)
```

```{r}
plot(out.dirch_hyper_model)
gelman.plot(out.dirch_hyper_model)
```

Bad convergence. Let's think about what we just did. We said that the proportion of each taxon at each location is drawn from the concentration parameters of each taxon at all locations. So, we are assuming that some taxa are more abundant than others across all locations, which may not be appropriate.

## Model 4: Site-specific parameters

While more complicated, it is probably worth making the $\alpha$ parameters site-specific:

$$\boldsymbol{y}_i \sim \text{Dirichlet}(\boldsymbol{\alpha}_i)$$

$$\alpha_{i,j} \sim \text{Gamma}(\alpha_{\alpha,j}, \beta_{\beta,j})$$

$$\alpha_{\alpha,j} \sim \text{Gamma}(0.001, 0.001)$$

$$\beta_{\alpha,j} \sim \text{Gamma}(0.001, 0.001)$$

Here, I have also chosen to allow the concentration of each taxon to be drawn independently from the uninformative prior. This was done to allow more flexibility in the model, so that the concentration hierarchically differs by site, then by taxon.

```{r}
dirch_hyper_unpool_model <- "
model{
for(i in 1:nobs){
y[i,1:ntaxa] ~ ddirch(alpha[i,1:ntaxa])
for(j in 1:ntaxa){
alpha[i,j] ~ dgamma(alpha_alpha[j], beta_alpha[j])
}
}
for(j in 1:ntaxa){
alpha_alpha[j] ~ dgamma(0.001, 0.001)
beta_alpha[j] ~ dgamma(0.001, 0.001)
}
}
"
```

```{r}
data <- list(nobs = nrow(data_sub),
             ntaxa = ncol(data_sub),
             y = data_sub)
jags.dirch_hyper_unpool_model <- jags.model(file = textConnection(dirch_hyper_unpool_model),
                                            data = data,
                                            n.chains = 3)
out.dirch_hyper_unpool_model <- coda.samples(model = jags.dirch_hyper_unpool_model,
                                             variable.names = c('alpha_alpha', 'beta_alpha'),
                                             n.iter = 1000)
```

```{r}
#plot(out.dirch_hyper_unpool_model)
#gelman.plot(out.dirch_hyper_unpool_model)
```

## Model 5: Incorporating climate drivers

Now, let's model the $\alpha$ parameters as a function of climate drivers:

$$\boldsymbol{y}_i \sim \text{Dirichlet}(\boldsymbol{\alpha}_i)$$

$$\alpha_{i,j} \sim \text{Gamma}(\alpha_{\alpha,i,j}, \beta_{\beta,i,j})$$

$$\alpha_{\alpha,i,j} = \frac{\mu_{i,j}^2}{\sigma_{i,j}^2}$$

$$\beta_{\alpha,i,j} = \frac{\mu_{i,j}}{\sigma_{i,j}^2}$$

$$\mu_{i,j} = \beta_0 + beta_{1,j}x_{i,1} + beta_{2,j}x_{i,2}$$

$$\sigma_{i,j}^2 = \frac{1}{\tau_{i,j}^2}$$

$$\tau_{i,j} \sim \text{Gamma}(0.001, 0.001)$$

$$\boldsymbol{\beta} \sim \text{MVN}(\mu_{\beta}, \tau_{\beta})$$

```{r}
dirch_driver_model <- "
model{
for(i in 1:nobs){
y[i,1:ntaxa] ~ ddirch(alpha[i,1:ntaxa])
for(j in 1:ntaxa){
alpha[i,j] ~ dgamma(alpha_alpha[i,j], beta_alpha[i,j])

alpha_alpha[i,j] <- mu[i,j]^2 / sigma.sq[j]
beta_alpha[i,j] <- mu[i,j] / sigma.sq[j]

logit(mu[i,j]) <- beta[1] + beta[match[1,j]] * x[i,1] + beta[match[2,j]] * x[i,2]
}
}
for(j in 1:ntaxa){
sigma.sq[j] <- 1 / tau[j]
tau[j] ~ dgamma(0.001, 0.001)
}
beta ~ dmnorm(mu_beta, tau_beta)
}
"
```

```{r}
maxparam <- (2 * ntaxa) + 1
match <- seq(from = 2, to = maxparam, by = 1)
match <- matrix(match, nrow = 2, ncol = ntaxa)

data <- list(nobs = nrow(data_sub),
             ntaxa = ncol(data_sub),
             y = data_sub,
             x = clim_sub,
             match = match,
             mu_beta = rep(x = 0, times = maxparam),
             tau_beta = diag(x = 0.001,
                             nrow = maxparam,
                             ncol = maxparam))
jags.dirch_driver_model <- jags.model(file = textConnection(dirch_driver_model),
                                      data = data,
                                      n.chains = 3)
out.dirch_driver_model <- coda.samples(model = jags.dirch_driver_model,
                                       variable.names = c('beta', 'tau'),
                                       n.iter = 1000)
```

While not converged, this is really promising given how few observations we are using right now. This is a good start, but we aren't just interested in the relationship between climate and vegetation. We're also interested in how the vegetation moderates that relationship, and how space and time factor in.

## Model 6: Incorporating species covariance

The next task is to add species covariance into the model. This is accomplished by modeling the $\alpha$ parameters of the Dirichlet model jointly using a multivariate normal distribution. The multivariate normal distribution was chosen because it is the only distribution that is implementable in JAGS that is appropriate for our data. Although a multivariate gamma distribution would be more appropriate, this is not available in JAGS and so was not used here. (It should be noted that there is a multivariate gamma distribution in JAGS, which is the Wishart distribution. However, the Wishart distribution describes the distribution of a matrix, not a vector, which is what we need here.)

In order to build the necessary functionality into the existing model, I retraced a couple of steps to incorporate the multivariate normal distribution jointly describing $\boldsymbol{\alpha}$. First, I modeled fractional composition from a Dirichlet distribution, with priors on the concentration $\boldsymbol{\alpha}$ parameter vector at all sites combined:

$$\boldsymbol{y}_i \sim \text{Dirichlet}(\boldsymbol{\alpha})$$

$$\boldsymbol{alpha} \sim \text{MVN}(\boldsymbol{\mu}, \boldsymbol{\Sigma})$$

where $\boldsymbol{\mu}$ is a vector of 0.5 of length $n_{\text{taxa}}$ and $\boldsymbol{\Sigma}$ is a diagonal $n_{\text{taxa}} \times n_{\text{taxa}}$ matrix with diagonal elements of 0.001 and all other elements = 0.

```{r}
mvn_model <- "
model{
for(i in 1:nobs){
y[i,1:ntaxa] ~ ddirch(alpha[1:ntaxa])
}
alpha[1:ntaxa] ~ dmnorm(mu, Sigma)
}
"
```

```{r}
nobs <- nrow(data_sub)
ntaxa <- ncol(data_sub)

data <- list(nobs = nobs,
             ntaxa = ntaxa,
             mu = rep(0.5, times = ntaxa),
             Sigma = diag(0.001,
                          nrow = ntaxa,
                          ncol = ntaxa))

jags.mvn_model <- jags.model(file = textConnection(mvn_model),
                             data = data,
                             n.chains = 3)

out.mvn_model <- coda.samples(model = jags.mvn_model,
                              variable.names = 'alpha',
                              n.iter = 1000)
```

Next, I incorporated hyperpriors on the distribution of the concentration parameters:

$$\boldsymbol{y}_i \sim \text{Dirichlet}(\boldsymbol{\alpha})$$

$$\boldsymbol{\alpha} \sim \text{MVN}(\boldsymbol{\mu}, \boldsymbol{\Sigma})$$

$$\boldsymbol{\mu} \sim \text{MVN}(\boldsymbol{\mu}_{\mu}, \boldsymbol{\Sigma}_{\mu})$$

$$\boldsymbol{\Sigma} \sim \text{Wishart}(\boldsymbol{R}, k)$$

The only difference between this model and the previous is the inclusion of hyper priors on $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}$. This is necessary because the objective of modeling $\boldsymbol{\alpha}$ jointly is to estimate the covariance between the taxa, as represented by the concentration parameters. This necessitates estimating $\boldsymbol{\Sigma}$. I assumed the mean, $\boldsymbol{\mu}$ to be normally distributed using a multivariate normal distribution, while for the covariance $\boldsymbol{\Sigma}$, I used the Wishart distribution, which is often used for specifying prior distributions ofr covariance matrices. I specified vague priors for the Wishart hyperprior, with $\boldsymbol{R}$ a diagonal $n_{\text{taxa}} \times n_{\text{taxa}}$ matrix with diagonal elements = 0.001 and off-diagonal elements = 0, and $k = n_{\text{taxa}} + 2$ because for the case of a vague prior, $k$ should be $> n$.

```{r}
mvn_hyper_model <- "
model{
for(i in 1:nobs){
y[i,1:ntaxa] ~ ddirch(alpha[1:ntaxa])
}
alpha[1:ntaxa] ~ dmnorm(mu, Sigma)
mu ~ dmnorm(mu_mu, Sigma_mu)
Sigma ~ dwish(R, k)
}
"
```

```{r}
data <- list(nobs = nobs,
             ntaxa = ntaxa,
             y = data_sub,
             mu_mu = rep(0.5, times = ntaxa),
             Sigma_mu = diag(0.001,
                             nrow = ntaxa,
                             ncol = ntaxa),
             R = diag(0.001,
                      nrow = ntaxa,
                      ncol = ntaxa),
             k = ntaxa + 2)

jags.mvn_hyper_model <- jags.model(file = textConnection(mvn_hyper_model),
                                   data = data,
                                   n.chains = 3, n.adapt = 2000)

out.mvn_hyper_model <- coda.samples(model = jags.mvn_hyper_model,
                                    variable.names = c('alpha', 'mu', 'Sigma'),
                                    n.iter = 1000)
```

Finally, I incorporate the climate drivers into the model with species covariance. This is done essentially by combining the model with climate drivers and the model of species covariance, both of which are detailed above:

$$\boldsymbol{y}_i \sim \text{Dirichlet}(\boldsymbol{\alpha})$$

$$\boldsymbol{\alpha}_i \sim \text{MVN}(\boldsymbol{\mu}_i, \boldsymbol{\Sigma})$$

$$\mu_{i,j} = exp(\beta_0 + \beta_{1,j} \times x_{i,1} + \beta_{2,j} \times \x_{i,2})$$

$$\boldsymbol{\Sigma} \sim \text{Wishart}(\boldsymbol{R}, k)$$

$$\boldsymbol{\beta} \sim \text{MVN}(\boldsymbol{\mu}_{\beta}, \boldsymbol{\Sigma}_{\beta})$$

```{r}
sp_clim_model <- "
model{
for(i in 1:nobs){
y[i,1:ntaxa] ~ ddirch(alpha[i,1:ntaxa])
alpha[i,1:ntaxa] ~ dmnorm(mu[i,1:ntaxa], Sigma)
for(j in 1:ntaxa){
logit(mu[i,j]) <- beta[1] + beta[match[1,j]] * x[i,1] + beta[match[2,j]] * x[i,2]
}
}
Sigma ~ dwish(R, k)
beta ~ dmnorm(mu_beta, Sigma_beta)
}
"
```

```{r}
data <- list(nobs = nobs,
             ntaxa = ntaxa,
             y = data_sub,
             match = match,
             x = clim_sub,
             R = diag(0.001,
                      nrow = ntaxa,
                      ncol = ntaxa),
             k = ntaxa + 4,
             mu_beta = rep(0, times = maxparam),
             Sigma_beta = diag(0.001,
                               nrow = maxparam,
                               ncol = maxparam))

jags.sp_clim_model <- jags.model(file = textConnection(sp_clim_model),
                                 data = data,
                                 n.chains = 3, n.adapt = 2000)

out.sp_clim_model <- coda.samples(model = jags.sp_clim_model,
                                  variable.names = c('beta', 'Sigma'),
                                  n.iter = 1000)
```